---
title: "Conditional vs prospective power analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditional vs prospective power analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## The problem

Most ecological monitoring programmes face periodic design reviews: should we
keep all plots, drop some, change the visit frequency, adjust survey effort?
These decisions require power analysis — but the standard tool for power
analysis of mixed models (`simr`) answers the wrong question.

`simr` performs **prospective** power analysis.  It initialises every plot
from visit zero, simulates a complete monitoring record from scratch, fits
the model, and checks whether the trend is significant.  This answers:

> "If we were starting a new programme today, how many plots would we need?"

But the actual question is:

> "Given what we've already collected, what happens if we change the design?"

These are different questions with different answers.

---

## Two modes of simulation

`monpwr` supports both modes explicitly.

### Prospective mode

Every plot starts at `visit_num = 0` with a linear predictor set to the
population marginal intercept.  Fresh random effects are drawn each
replicate from $N(0, \hat\sigma)$.  No historical data enters the trend
model — only simulated future visits.

This is methodologically equivalent to `simr`.  It answers the
"starting from scratch" question.

### Conditional mode

Each plot is initialised at its **last observed visit number** with a
linear predictor derived from the fitted model — absorbing all fixed
covariate effects and the empirical BLUP.  Future visits are projected
forward from this position.  The historical data stub (real observed data)
is stacked with simulated future data, so the trend model sees the
full temporal record.

This answers the "given what we have" question.

---

## Why the distinction matters

### Temporal capital

A plot that has been visited 5 times already has accumulated information
about its trajectory.  When the trend model sees 5 real visits plus 5
simulated future visits, it has 10 data points per plot.  Prospective
simulation gives it only 5 (all simulated).  The additional leverage from
banked visits is "temporal capital" — monitoring investment that
prospective analysis discards.

The practical consequence: prospective power analysis **systematically
underestimates** the power of established monitoring programmes.  Design
reviews that rely on prospective analysis may conclude that a programme
lacks adequate power when it actually doesn't, leading to unnecessarily
conservative (and expensive) design recommendations.

### Plot-specific initialisation

Conditional mode initialises each plot individually.  This means:

- Plots with different monitoring histories start at different positions
- The `site_selector` in each scenario picks **specific real plots**, not
  a generic sample size
- Two scenarios with the same *n* can give different power estimates
  depending on which plots are retained

`simr` cannot represent any of this.  It works with a single data template
where every plot is interchangeable.

### BLUPs as point estimates

Conditional mode uses the empirical BLUP (best linear unbiased predictor)
as each plot's starting point.  This is a deliberate choice: sampling from
the BLUP's posterior variance would introduce additional uncertainty that
shifts conditional simulation toward prospective, undermining the framing.

Shrinkage of BLUPs toward the population mean is a caveat to acknowledge
in interpretation, not a problem to correct architecturally.

---

## Running both modes

The core workflow for a design review is to run both modes on the same
data and compare:

```{r comparison}
library(monpwr)

# Extract parameters from a fitted model
ref <- extract_params(fit, data = as.data.frame(survey_data),
                      visit_num_var = "visit_num",
                      plotid_var    = "site_id",
                      place_var     = "site_id")

scenarios <- list(
  full    = scenario("All sites",  1, function(sp) unique(sp$site_id)),
  reduced = scenario("Core only",  1, function(sp) sp$site_id[sp$is_core])
)

# Conditional: what power do we actually have?
results_cond <- run_power_sim(
  ref, scenarios,
  plot_metadata    = site_metadata,
  mode             = "conditional",
  data             = as.data.frame(survey_data),
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  n_iter           = 500
)

# Prospective: what would simr tell us?
results_prosp <- run_power_sim(
  ref, scenarios,
  plot_metadata    = site_metadata,
  mode             = "prospective",
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  n_iter           = 500
)
```

The gap between `results_cond` and `results_prosp` is the temporal capital
that prospective analysis ignores.  For programmes with 5+ years of data,
this gap can be substantial.

---

## When to use each mode

| Question | Mode |
|----------|------|
| "How many plots do we need for a new programme?" | Prospective |
| "Should we keep 100 plots or drop to 60?" | Conditional |
| "What is our current power to detect a 20% decline?" | Conditional |
| "Would adding 20 new sites improve power?" | Conditional (with hybrid scenarios) |
| "How does our design compare to starting from scratch?" | Both — the gap is informative |

---

## The conditional advantage is not just statistical

The gap between conditional and prospective power estimates has direct
funding implications.  Monitoring programmes compete for limited
conservation budgets.  If a prospective analysis says 80 plots are needed
to detect a 20% decline, but conditional analysis shows that 50 of your
existing plots — with their accumulated histories — already achieve 80%
power, that's 30 plots' worth of fieldwork budget that can be redirected
elsewhere.

Conversely, conditional analysis can reveal that specific plots contribute
disproportionately to power (because they have long histories or are in
informative parts of the parameter space), guiding decisions about which
plots to prioritise when budgets contract.

---

## Relationship to simr

`monpwr`'s prospective mode is methodologically equivalent to `simr`.
Both initialise plots from scratch, draw fresh random effects, and test
for a trend.  The differences are:

1. **Model support**: `simr` requires `lme4` models; `monpwr` supports
   `glmmTMB` (hurdle models, zero-inflation, nbinom2) and `lme4`.

2. **Test model**: `simr` re-fits the original model; `monpwr` fits a
   simplified test model (random intercept only).  This is deliberately
   conservative.

3. **Design-space exploration**: `monpwr`'s scenario/grid architecture
   makes it natural to sweep over sample sizes, effort levels, horizons,
   and effect sizes simultaneously.

4. **Retest and extend**: `monpwr` stores raw p-values, enabling
   `retest()` at alternative alpha levels and `extend()` to pool
   additional iterations — both without re-simulating.

5. **Conditional mode**: `simr` cannot do this.  Its architecture assumes
   a fixed data template; `monpwr`'s conditional mode changes the data
   frame each replicate based on which plots are retained and their
   individual histories.

For a validation exercise, running `monpwr` prospective alongside `simr`
on the same model and parameters should produce statistically
indistinguishable power estimates.  The conditional mode is then the
contribution that `simr` cannot replicate.
