---
title: "Getting started with monpwr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with monpwr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set to TRUE once real data objects are available
)
```

## Overview

`monpwr` evaluates statistical power and precision for biodiversity monitoring
programme design scenarios.  It supports two complementary simulation modes:

- **Conditional** — plots are initialised at their observed visit history and
  BLUP from a fitted model. Future visits are projected forward, so the
  existing temporal record contributes to power. This is the right tool when
  you already have monitoring data and are deciding which plots to continue
  visiting.

- **Prospective** — plots are initialised from scratch using model parameters
  only (no existing record). Equivalent to `simr`-style analysis. Use this
  when evaluating a brand-new design or comparing "what power would we have
  had if we'd used design X from the start".

Both modes work with any model that `extract_params()` supports — currently
`glmmTMB` and `lme4` model objects.

---

## 1. Fit your model

```{r fit}
library(monpwr)
library(glmmTMB)

# Example: hurdle model for pellet count data
# Replace with your own model and dataset
fit <- glmmTMB(
  count ~ visit_num + offset(log_effort) + (1 | plotid_model),
  ziformula = ~ (1 | plotid_model),
  family    = truncated_nbinom2,
  data      = long_model   # your data frame — must be a plain data.frame
)
```

---

## 2. Extract standardised parameters

`extract_params()` dispatches on the model class and returns a `monpwr_params`
object containing everything the simulation engine needs.

```{r extract}
# data must be a plain data.frame — call as.data.frame() first if needed
long_model <- as.data.frame(long_model)

ref <- extract_params(fit, data = long_model)
ref
```

The returned object has a fixed structure regardless of model family, so all
downstream functions are model-agnostic.

---

## 3. Define design scenarios

Each scenario specifies a remeasurement interval and a `site_selector` —
a function that takes your plot-metadata data frame and returns a vector of
plot IDs to retain.

```{r scenarios}
scenarios <- list(

  baseline = scenario(
    label         = "Full grid, 5-yr",
    remeasure_yrs = 5,
    site_selector = function(sp) unique(sp$Place)
  ),

  coarse = scenario(
    label         = "Coarse grid, 5-yr",
    remeasure_yrs = 5,
    site_selector = function(sp) sp$Place[sp$in_coarse_grid]
  ),

  extended = scenario(
    label         = "Full grid, 10-yr",
    remeasure_yrs = 10,
    site_selector = function(sp) unique(sp$Place)
  )
)
```

---

## 4. Run simulations

### Conditional mode (recommended when you have existing data)

```{r run-conditional}
results <- run_power_sim(
  ref_params       = ref,
  scenarios        = scenarios,
  plot_metadata    = fpi_spatial,   # one row per plot with Place, Park_type, etc.
  mode             = "conditional",
  data             = long_model,    # required for conditional mode
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  n_iter           = 200,
  alpha            = 0.10,
  scales           = c("National", "Park type"),
  park_type_var    = "Park_type"
)
```

### Prospective mode (from-scratch comparison)

```{r run-prospective}
results_p <- run_power_sim(
  ref_params       = ref,
  scenarios        = scenarios,
  plot_metadata    = fpi_spatial,
  mode             = "prospective",  # no `data` argument needed
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  n_iter           = 200
)
```

---

## 5. Summarise and plot

```{r summarise}
# Minimum detectable change
mdc <- compute_mdc(results, power_target = 0.80)

# Power curve
plot_power(results, scale_filter = "National")

# MDC heatmap
plot_mdc(mdc, scale_filter = "National")

# Precision (CV) of state estimates
precision <- run_precision(scenarios, data = long_model,
                           plot_metadata = fpi_spatial)
plot_cv(precision, scale_filter = "National")
```

---

## 6. Custom model families

If your model is not `glmmTMB` or `lme4`, implement an `extract_params` S3
method for your model class:

```{r custom-extractor}
extract_params.my_model_class <- function(fit, data,
                                          visit_num_var = "visit_num",
                                          plotid_var    = "plotid_model",
                                          place_var     = "Place",
                                          ...) {
  # Extract parameters from your model
  # ...

  # Return a list with class "monpwr_params"
  # Required fields: beta_visit, beta_gap_cond, beta_gap_zi, disp_par,
  #                  sigma_cond, sigma_zi, visit_gap_med, family,
  #                  visit_num_var, plotid_var, place_var, plot_state
  structure(
    list(
      beta_visit    = ...,
      beta_gap_cond = 0,
      beta_gap_zi   = 0,
      disp_par      = 1,
      sigma_cond    = ...,
      sigma_zi      = 0,
      visit_gap_med = median(data$visit_gap, na.rm = TRUE),
      family        = "poisson",   # or whichever family string matches
      visit_num_var = visit_num_var,
      plotid_var    = plotid_var,
      place_var     = place_var,
      plot_state    = ...          # data.frame: Place, plotid_model, visit_num,
                                   #             eta_last_cond, eta_last_zi,
                                   #             blup_cond, blup_zi
    ),
    class = "monpwr_params"
  )
}
```

---

## 7. Custom prospective initialisers

The default prospective initialiser (`init_prospective_marginal`) sets all
plots to the population-average linear predictor.  If you want plots
initialised with specific covariate profiles, supply your own `init_fn`:

```{r custom-init}
# Example: sample covariate profiles from the empirical data
my_init_fn <- function(ref_params, n_plots, template_data = NULL, ...) {
  # template_data: a data frame of plot-level covariates to sample from
  sampled <- template_data[sample(nrow(template_data), n_plots, replace = TRUE), ]

  data.frame(
    plotid        = paste0("plot_", seq_len(n_plots)),
    visit_num     = 0L,
    eta_last_cond = sampled$my_eta_cond,  # your pre-computed LPs
    eta_last_zi   = sampled$my_eta_zi
  )
}

results_custom <- run_power_sim(
  ref, scenarios,
  plot_metadata    = fpi_spatial,
  mode             = "prospective",
  init_fn          = my_init_fn,
  template_data    = fpi_spatial,   # passed through via ...
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20)
)
```
