---
title: "Getting started with monpwr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with monpwr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`monpwr` estimates statistical power and minimum detectable change (MDC) for
biodiversity monitoring programme design.  It supports two simulation modes:

- **Conditional** — plots initialised at their observed visit history and BLUP.
  The existing temporal record contributes to the trend estimate.  Use this when
  deciding which plots to retain in an *existing* programme.

- **Prospective** — plots initialised from scratch using model parameters only.
  Equivalent to `simr`-style analysis.  Use for new programme design or as a
  comparison baseline.

Both modes work with any model that `extract_params()` supports — currently
`glmmTMB` and `lme4` model objects.

---

## 1. Fit your model

`monpwr` extracts parameters from a fitted mixed model.  The model should
include a visit-sequence variable as a fixed effect and a plot-level random
intercept.

```{r fit}
library(monpwr)
library(glmmTMB)

# Example: hurdle NB2 model for species counts
# Replace with your own model and dataset
fit <- glmmTMB(
  count ~ visit_num + offset(log_effort) + (1 | site_id),
  ziformula = ~ (1 | site_id),
  family    = truncated_nbinom2,
  data      = survey_data   # must be a plain data.frame, not a tibble
)
```

**Important:** `data` must be a plain `data.frame` — not a tibble or
`data.table`.  BLUP extraction via `ranef()` can silently fail otherwise.
Call `as.data.frame()` before passing to the package.

---

## 2. Extract standardised parameters

`extract_params()` dispatches on model class and returns a `monpwr_params`
object — a standardised list containing everything the simulation engine needs.

```{r extract}
ref <- extract_params(
  fit,
  data          = as.data.frame(survey_data),
  visit_num_var = "visit_num",   # column name of visit sequence variable
  plotid_var    = "site_id",     # column name of RE grouping factor
  place_var     = "site_id"     # column name of plot identifier
)
ref
```

Column name arguments (`visit_num_var`, `plotid_var`, `place_var`) map your
dataset's column names to monpwr's internal conventions.  If your RE grouping
factor and plot identifier are the same column, pass the same name for both.

### Offset handling

If your model includes an offset, tell `extract_params()` how to find it:

```{r extract-offset}
# Option 1: pre-computed log-offset column in data
ref <- extract_params(fit, data = dat, offset_var = "log_effort")

# Option 2: inline offset transformation
ref <- extract_params(fit, data = dat,
                      offset_transform = function(d) log(d$n_hours))

# Override the assumed effort for future simulated visits (default = median observed)
ref <- extract_params(fit, data = dat,
                      offset_transform = function(d) log(d$n_hours),
                      log_effort_future = log(4))
```

---

## 3. Define design scenarios

Each scenario specifies a remeasurement interval and a `site_selector` —
a function that takes your plot metadata and returns the plot IDs to retain.

```{r scenarios}
scenarios <- list(

  full = scenario(
    label         = "All sites, annual",
    remeasure_yrs = 1,
    site_selector = function(sp) unique(sp$site_id)
  ),

  reduced = scenario(
    label         = "Core sites only, annual",
    remeasure_yrs = 1,
    site_selector = function(sp) sp$site_id[sp$is_core]
  ),

  biennial = scenario(
    label         = "All sites, biennial",
    remeasure_yrs = 2,
    site_selector = function(sp) unique(sp$site_id)
  )
)
```

---

## 4. Run simulations

### Conditional mode (when you have existing data)

```{r run-conditional}
results <- run_power_sim(
  ref_params       = ref,
  scenarios        = scenarios,
  plot_metadata    = site_metadata,   # one row per plot
  mode             = "conditional",
  data             = as.data.frame(survey_data),
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  n_iter           = 200,
  alpha            = 0.10
)
```

### Prospective mode (new programme design)

```{r run-prospective}
results_p <- run_power_sim(
  ref_params       = ref,
  scenarios        = scenarios,
  plot_metadata    = site_metadata,
  mode             = "prospective",
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  n_iter           = 200
)
```

### Sub-group breakdowns

To stratify results by a grouping variable (e.g. region, habitat type),
pass `group_var`:

```{r run-grouped}
results_grouped <- run_power_sim(
  ref_params       = ref,
  scenarios        = scenarios,
  plot_metadata    = site_metadata,
  mode             = "conditional",
  data             = as.data.frame(survey_data),
  effect_sizes_pct = c(10, 20, 30),
  horizons         = c(10, 20),
  group_var        = "region"   # column in site_metadata
)
```

This produces one "All" row (all selected sites) plus one row per unique
level of `region`.

---

## 5. Explore results

```{r explore}
# Print summary
print(results)

# Check convergence
results |>
  dplyr::filter(conv_rate < 0.90) |>
  dplyr::select(label, effect_pct, horizon, power, conv_rate)

# Power bar chart
plot_power(results)

# Minimum detectable change
mdc <- compute_mdc(results, power_target = 0.80)
plot_mdc(mdc)
```

---

## 6. Varying sample size and effort

A common design question is "how many plots at what effort level?"
Loop over combinations outside `run_power_sim()`:

```{r design-grid}
library(purrr)
library(tidyr)

grid <- expand_grid(
  n_plots    = c(10, 20, 30, 50, 80),
  effort_hrs = c(2, 4, 6, 8)
)

results_grid <- pmap_dfr(grid, function(n_plots, effort_hrs) {
  rp <- ref
  rp$log_effort_future <- log(effort_hrs)

  sc <- list(
    sc1 = scenario(
      label         = paste0(n_plots, " plots, ", effort_hrs, " hrs"),
      remeasure_yrs = 1,
      site_selector = function(sp) head(sp$site_id, n_plots)
    )
  )

  run_power_sim(
    scenarios        = sc,
    ref_params       = rp,
    plot_metadata    = site_metadata,
    mode             = "prospective",
    horizons         = 10,
    effect_sizes_pct = c(5, 10, 15, 20, 30),
    n_iter           = 200
  ) |>
    dplyr::mutate(n_target = n_plots, effort_hrs = effort_hrs)
})
```

Visualise with `plot_power_grid()`:

```{r power-grid}
plot_power_grid(
  results_grid,
  y_var     = "n_target",
  x_var     = "effort_hrs",
  facet_var = "effect_pct",
  y_lab     = "n plots",
  x_lab     = "Survey effort (hrs)",
  title     = "Power by sample size and effort — 10 yr horizon"
)
```

---

## 7. Retest and extend

`monpwr` stores every replicate's raw p-value, enabling two workflows
that avoid re-simulating:

### Retest at a different alpha

```{r retest}
# Original run used alpha = 0.10; what about alpha = 0.05?
results_05 <- retest(results, alpha = 0.05)
```

### Extend with additional iterations

If some cells have wide confidence intervals, run more iterations on just
those cells and pool with the original:

```{r extend}
# Identify cells with wide CIs
wide <- results |>
  dplyr::mutate(ci_width = power_upper - power_lower) |>
  dplyr::filter(ci_width > 0.10)

# Run additional iterations for those cells, then pool
results_topup <- run_power_sim(...)  # same parameters, more n_iter
results_pooled <- extend(results, results_topup)
```

---

## 8. Custom model families

If your model is not `glmmTMB` or `lme4`, implement an `extract_params`
S3 method for your model class:

```{r custom-extractor}
extract_params.my_model_class <- function(fit, data,
                                          visit_num_var = "visit_num",
                                          plotid_var    = "site_id",
                                          place_var     = "site_id",
                                          ...) {
  # Extract your model's parameters and return a monpwr_params object.
  # All fields listed below are required.
  structure(
    list(
      beta_visit        = ...,
      beta_gap_cond     = 0,
      beta_gap_zi       = 0,
      disp_par          = ...,
      sigma_cond        = ...,
      sigma_zi          = 0,
      visit_gap_med     = 0,
      family            = "poisson",
      visit_num_var     = visit_num_var,
      plotid_var        = plotid_var,
      place_var         = place_var,
      visit_gap_var     = "visit_gap",
      count_var         = "count",
      offset_var        = NULL,
      log_effort_future = 0,
      plot_state        = ...   # data.frame with columns: place_id, plotid,
                                # visit_num, eta_last_cond, eta_last_zi,
                                # blup_cond, blup_zi
    ),
    class = "monpwr_params"
  )
}
```
